Free surface computations in mantle convection models

### 2016-04-04 free-surface-paper.tex
Use line numbering
Make equation references consistent.

@@

In order to better understand the time evolution of the system in \eqref{eq:final_system}
we will consider the eigenvalues of a simplified homogeneous system. By setting $\rho = \rho_0$, we obtain the system
\begin{equation}
\begin{aligned}
\nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \nabla p^\prime &= 0 \\
\nabla \cdot \mathbf{u} &= 0
\end{aligned}
\label{eq:homogeneous_stokes}
\end{equation}

@@

Armed with the normal modes and relaxation times of the Stokes system, we can write down the
formal solution to Equation~\ref{eq:surface_evolution}. Let the initial surface topography be 
represented by a linear combination of its normal modes
\begin{equation}
\zeta(\mathbf{x}, t=0) = \displaystyle \sum_i a_i \zeta_i(\mathbf{x})
\end{equation}
then the time evolution of the free surface is given by
\begin{equation}
\zeta( \mathbf{x}, t) = \displaystyle \sum_i a_i \zeta_i(\mathbf{x}) e^{-t/\tau_i}.
\end{equation}
Of course, most finite element (or finite difference, or finite volume) geodynamic simulations do not resolve 
the solution and surface topography into its normal modes and integrate those separately. 
Indeed, analytical solutions for the normal modes are only known for simple geometries and rheologies.
Instead, they obtain a velocity solution and simply advect the free surface with the local velocity.
The normal mode solution is instructive, however. 
Each mode has the form of a decay equation with characteristic decay time $\tau_i$.
The decay equation is the archetypical example of a stiff ordinary differential equation.
If we were to numerically integrate this in time with a forward Euler method, we would find the 
time-step criterion for stability \citep[e.g.][]{leveque2007finite} to be
\begin{equation}
\Delta t  \le 2 \tau_{\mathrm{min}},
\label{eq:cfl_euler}
\end{equation}
where $\tau_{\mathrm{min}}$ is this minumum decay time.
The maximum allowable timestep is limited by the minimum relaxation timescale.
If a larger timestep than this is taken then those modes will go unstable.
The modes with the smallest relaxation times are usually those with the largest lengthscales \citep{schubert2001mantle}, 
and so it will be those which go unstable first, a phenomenon which has been called 
a ``sloshing'', or ``drunken sailor'' instability \citep{kaus2010stabilization}.

@@

The stability of this scheme is determined by the choice of $\tau^*$.  
It has a region of unconditional stability, where
\begin{equation}
\tau^* \le 2 \tau_i.
\end{equation}
Otherwise, the scheme is conditionally stable, with the timestep limited by 
\begin{equation}
\Delta t \le -\tau^* \log \left(1 - 2 \frac{\tau_i}{\tau^*} \right).
\end{equation}
The stability region is plotted in Figure \ref{fig:stability_region}.

@@

\begin{itemize}
\item Direct solution of Equation~\eqref{eq:generalized_eigenvalue}. This can be expensive, although
only the minimun relaxation time is required. For certain geometries a simple power iteration on 
the standard eigenvalue problem~\eqref{eq:standard_eigenvalue} could be enough.
\item Analytical formulae. Several geometry and viscosity model combinations have analytical solutions
for relaxation spectra. Even if the user's model is not precisely the same (e.g., has some lateral viscosity
variations), an analytical approximation may be sufficient.
\item Scaling. In general, we expect the relaxation times to scale with $\tau \sim \frac{\eta}{\rho g L}$,
where the density, gravity, viscosity, and lengthscale $L$ are all representative values.  A small amount
of experimentation near this value of $\tau$ can find an appropriate relaxation time.
\item Observation of instabilities.  The mere act of observing a sloshing instability in an unstabilized
problem can furnish an estimate of its relaxation time.
\end{itemize}

@@

A better approach is to perform a finite element projection of the Stokes velocity 
solution onto the mesh velocity vector. Multiplying the boundary conditions 
\eqref{eq:laplacian_bcs} by a test function $\mathbf{w}$ and integrating over the free
surface part of the boundary, we find

@@

This weak solution to the boundary conditions~\eqref{eq:laplacian_bcs} is able to borrow
the accuracy of the Stokes solve for Equation~\eqref{eq:weak_incompressible}, and we have 
found it to conserve mass more accurately than algebraic techniques for evaluating the mesh-normal velocity.
Similar results were found by \citet{fullsack1995arbitrary}.

>>>>

\usepackage{lineno}
    \linenumbers

@@

In order to better understand the time evolution of the system in Equation~\eqref{eq:final_system}
we will consider the eigenvalues of a simplified homogeneous system. By setting $\rho = \rho_0$, we obtain the system
\begin{equation}
\begin{aligned}
\nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \nabla p^\prime &= 0 \\
\nabla \cdot \mathbf{u} &= 0
\end{aligned}
\label{eq:homogeneous_stokes}
\end{equation}

@@

Armed with the normal modes and relaxation times of the Stokes system, we can write down the
formal solution to Equation~\eqref{eq:surface_evolution}. Let the initial surface topography be 
represented by a linear combination of its normal modes
\begin{equation}
\zeta(\mathbf{x}, t=0) = \displaystyle \sum_i a_i \zeta_i(\mathbf{x})
\end{equation}
then the time evolution of the free surface is given by
\begin{equation}
\zeta( \mathbf{x}, t) = \displaystyle \sum_i a_i \zeta_i(\mathbf{x}) e^{-t/\tau_i}.
\end{equation}
Of course, most finite element (or finite difference, or finite volume) geodynamic simulations do not resolve 
the solution and surface topography into its normal modes and integrate those separately. 
Indeed, analytical solutions for the normal modes are only known for simple geometries and rheologies.
Instead, they obtain a velocity solution and simply advect the free surface with the local velocity.
The normal mode solution is instructive, however. 
Each mode has the form of a decay equation with characteristic decay time $\tau_i$.
The decay equation is the archetypical example of a stiff ordinary differential equation.
If we were to numerically integrate this in time with a forward Euler method, we would find the 
time-step criterion for stability \citep[e.g.][]{leveque2007finite} to be
\begin{equation}
\Delta t  \le 2 \tau_{\mathrm{min}},
\label{eq:cfl_euler}
\end{equation}
where $\tau_{\mathrm{min}}$ is this minumum decay time.
The maximum allowable timestep is limited by the minimum relaxation timescale.
If a larger timestep than this is taken then those modes will go unstable.
The modes with the smallest relaxation times are usually those with the largest lengthscales \citep{schubert2001mantle}, 
and so it will be those which go unstable first, a phenomenon which has been called 
a ``sloshing'', or ``drunken sailor'' instability \citep{kaus2010stabilization}.

@@

The stability of this scheme is determined by the choice of $\tau^*$.  
It has a region of unconditional stability, where
\begin{equation}
\tau^* \le 2 \tau_i.
\end{equation}
Otherwise, the scheme is conditionally stable, with the timestep limited by 
\begin{equation}
\Delta t \le -\tau^* \log \left(1 - 2 \frac{\tau_i}{\tau^*} \right).
\end{equation}
The stability region is plotted in Figure~\ref{fig:stability_region}.

@@

\begin{itemize}
\item Direct solution of Equation~\eqref{eq:generalized_eigenvalue}. This can be expensive, although
only the minimun relaxation time is required. For certain geometries a simple power iteration on 
the standard eigenvalue problem (Equation~\eqref{eq:standard_eigenvalue}) could be enough.
\item Analytical formulae. Several geometry and viscosity model combinations have analytical solutions
for relaxation spectra. Even if the user's model is not precisely the same (e.g., has some lateral viscosity
variations), an analytical approximation may be sufficient.
\item Scaling. In general, we expect the relaxation times to scale with $\tau \sim \frac{\eta}{\rho g L}$,
where the density, gravity, viscosity, and lengthscale $L$ are all representative values.  A small amount
of experimentation near this value of $\tau$ can find an appropriate relaxation time.
\item Observation of instabilities.  The mere act of observing a sloshing instability in an unstabilized
problem can furnish an estimate of its relaxation time.
\end{itemize}

@@

A better approach is to perform a finite element projection of the Stokes velocity 
solution onto the mesh velocity vector. Multiplying the boundary conditions 
(Equation~\eqref{eq:laplacian_bcs}) by a test function $\mathbf{w}$ and integrating over the free
surface part of the boundary, we find

@@

This weak solution to the boundary conditions (Equation~\eqref{eq:laplacian_bcs}) is able to borrow
the accuracy of the Stokes solve for Equation~\eqref{eq:weak_incompressible}, and we have 
found it to conserve mass more accurately than algebraic techniques for evaluating the mesh-normal velocity.
Similar results were found by \citet{fullsack1995arbitrary}.
